The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-09-14 paper.qmd
Major changes: adding line breaks at each sentence - makes it easier for git to track changes without merge issues. Added a citation to Vanderplas 2017, in part because it's important to say that lineups can be used for hypothesis testing, but they can also be used to compare visual power of different charts for actually seeing differences in data.
Minor changes: Fixed a few typos and grammar issues.

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
  - I am not sure when I need a citation and when I dont. Sometimes I could literally cite hundreds of papers to make a point, so I'm not sure if i should cite any at all.
- examples that aren't maps? Sure are a lot of maps.
-->

@@

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In the face of this overflow of information, tools that effectively summarize information down into simple and clear ideas become more valuable.

@@

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. @fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. Today, there is an abundance of publications on the topic which makes it is timely to construct a review of the field. In fact, there have already been several reviews published.

@@

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. Finally, we discuss how uncertainty visualisations can be effectively evaluated. When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling tha purpose.

## Spatial example
It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot. The choropleth map. Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey. We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach. These examples will need

@@

## Signal-supression
Uncertainty visualisation is required for transparency. The two choropleth maps that appear to be identical in @fig-choropleth highlights the issues with simply electing to ignore uncertainty. This sentiment appears frequently in the uncertainty visualisation literature. Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

Imagine you are trying to decide if you want to bring an umbrella with you to work. An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. Therefore, your decision will be improved with the inclusion of uncertainty. This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. In this sense, uncertainty is no more "special" to decision making than weight special to a BMI calculation.

@@

When we use "uncertainty visualization" to refer to graphics that simply communicate a variance or probability, we are classifying  visualisations by the data they display, not their visual features. Graphics, just like statistics, are not defined by their input data. A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special. Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019]. While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty. Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This is a reasonable approach. This is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. However, it is unclear if this approach is will for our purposes.

@@

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. Positivists abhore the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.

@@

### Visualising the "single integrated uncertain value"
Typically, when making visualisations, we want the visual channels to be separable, that is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel". Additionally, visualizing uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.

This means effective uncertainty visualisation should be leveraging integrability. That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020]. When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012]. Unfortunately relying on integrability may not give us the amount of control we want over our signal-supresison. Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals. Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].

@@

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another. There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals. If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. Unfortunately it is unclear how reliably this would work.

### There is no uncertainty in EDA
Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis which means there is no inference, and by extension, there is no uncertainty.

Some authors recognize inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.

@@

Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color. We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". To answer this question we do a series of t-tests on these estimates. All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other. The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. We cannot perform signal-suppression on both these signals simultaneously.

This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together. Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly. That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value. At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain.

@@

This does not mean that visualizing raw data will always prevent insignificant signal from getting through. @Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. However just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked.

@@

### Value extraction of uncertainty statistics
Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and the variance [@Kinkeldey2014]. This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition. By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainties ability to suppress statistics. This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.

@@

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018]. Science communication should be primarily concerned with accuracy.

@@

Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019]. Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. Risk aversion is considered irrational behaviour because it is a deviation from the behavior of a rational agent. Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational]. However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. Rational agents *by definition* should *ignore* uncertainty information. Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information. The only cases where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.

@@

The first method is what we call the "vague question" approach. The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. The authors then go on to compare the readers answer to a very specific ground truth. This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributiosn (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group furthest to the right". The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution. Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. These studies frequently expect participants to give deterministic answers for probabilistic questions. For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and aksed them "how likely is candidate B to win the election?". Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. Interestingly it is unclear how the participants were supposed to answer this question. @Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. In the first study, participants were required assume to the storm was equally intense reguardless of the probability of the probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage) but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

>>>>

<!-- TODO 
- convert to American spelling 
  (Susan) Is this necessary? I'm not sure it matters as long as you're consistent...
- May not need to reference the same paper again and again
  - I am not sure when I need a citation and when I dont. Sometimes I could literally cite hundreds of papers to make a point, so I'm not sure if i should cite any at all.
- examples that aren't maps? Sure are a lot of maps.
-->

@@

From entertainment choices to news articles to insurance plans, the modern citizen is so over-run with information in every aspect of their life it can be overwhelming. 
In the face of this overflow of information, tools that effectively reduce piles of information to simple and clear ideas become more valuable.

@@

Uncertainty visualisation is a relatively new field in research. 
Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite the new terminology visualisation of uncertainty has been present since the earliest times. 
For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field. 
In fact, there have already been several reviews published.

<!-- Susan: I think there's a missed opportunity here to reference the premise you started with, that when there are heaps of information, it's valuable to distill it down into simple facts -->

@@

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. 
First, we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. 
Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. 
Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. 
Fourth, we discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. 
Finally, we discuss how uncertainty visualisations can be effectively evaluated.
When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling that purpose.

## Spatial example

It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot, the choropleth map. 
Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey. 
We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach. 
These examples will need XXX incomplete sentence!! XXX

@@

## Signal-supression
Uncertainty visualisation is required for transparency. 
The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
This sentiment appears frequently in the uncertainty visualisation literature.
Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. 
Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

Imagine you are trying to decide if you want to bring an umbrella with you to work. 
An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. 
Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. 
Therefore, your decision will be improved with the inclusion of uncertainty. 
This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. 
In this sense, uncertainty is no more "special" to decision making than weight is special to a BMI calculation.

@@

When we use "uncertainty visualization" to refer to graphics that simply communicate a variance or probability, we are classifying visualisations by the data they display, not their visual features. 
Graphics, just like statistics, are not defined by their input data. 
A scatter plot that compares mean and a scatter plot that compares variances are both scatter plots. 
Given that there is no special class of visualisation for *other* statistics (such the median or maximum) there is no reason to assume visualisations that simply depict a variance, error, or probability to be special. 
Some authors implicitly suggest that that visualisations of variance or probability are differentiated due to the psychological heuristics involved in interpreting uncertainty [@Hullman2019].
While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017], there is no evidence that this psychological effect translates to issues with the visual representation of uncertainty. 
Again, given that we do not make these same visual considerations for other variables that elicit distaste or irrational behavior, there is no reason to assume this is what makes uncertainty visualisation so special.

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
You may think "well, I will just add a dimension to my plot that includes uncertainty". 
This is a reasonable approach. 
The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
However, it is unclear if this approach is 
sufficient XXX Word choice??? XXX
<!-- I changed will to sufficient, but I'm not sure that's what you were going for. -->
for our purposes.

@@

Descriptive statistics describe our sample as it is and summarizes large data down into an easy to swallow format. 
Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. 
In 19th century England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson). 
Practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations [@Otsuka2023]. 
In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature", that is, we need to assume that unobserved phenomena should be similar to observed phenomena [@Otsuka2023]. 
Positivists abhor the assumption of the "uniformity of nature" as they believed referencing the unobservable is bad science. 
In other words, these scientists embraces descriptive statistics and shunned inferential statistics due to the inherent uncertainty that came with them. 
Since uncertainty is non-existent in descriptive statistics, it is clear that uncertainty is a by-product of inference.
<!-- Inferential statistics didn't yet exist when positivists were dominant - they weren't making a choice between description and inference, as the choice didn't exist. This is slightly misleading as written, even if I generally agree with the concept you're trying to communicate. -->

@@

<!-- You've promised to describe what uncertainty is not... did you achieve that goal? If not, you probably should revisit how you're introducing the philosophical bit about descriptive statistics and positivism.  -->

### Visualising the "single integrated uncertain value"
Typically, when making visualisations, we want the visual channels to be separable. 
That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a "single integrated channel". 
Additionally, visualizing uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.

This means effective uncertainty visualisation should be leveraging integrability. 
That is, the visual channels of the uncertainty and the signal would need to be separately manipulable, but read as a single channel by the human brain. 
While most visual aesthetics *are* separable, there are some variables that have been shown to be integrable, such as color hue and brightness [@Vanderplas2020].
When visualizing uncertainty using its own visual channel, we can also consider visual semiotics and make sure to map uncertainty to intuitive visual channels, such as mapping more uncertain values to lighter colors [@Maceachren2012].
Unfortunately relying on integrability may not give us the amount of control we want over our signal-supression. 
Without a strong understanding of how these visual channels collapse down into a single channel, relying on integrability could create unintended consequences such as displaying phantom signals or hiding justified signals. 
Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015].

@@

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals. 
If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. 
In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. 
However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. 
For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. 
Unfortunately, it is unclear how reliably this would work.

### There is no uncertainty in EDA
Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.

Some authors recognize inference will always occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. 
@Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. 
However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. 
<!-- Ok, so this is an additional challenge on top of Hullman's claim, but it's not refuting Hullman's claim. I think the easier way to do that is to say that while you can certainly think of visualizations as containing implicit inferential properties, there are many potential inferences in any single visualization; until the viewer chooses one and thinks about it (which isn't easily observable), the particular variety of uncertainty which would need to be displayed can't be calculated - it's a quantum physics type issue where until something is observed, the singularity doesn't collapse and everything exists only in probabilistic terms. The cat isn't dead or alive until the viewer starts thinking inferentially in one way or another. -->
This means an ideal uncertainty visualisations should not only meet the signal suppression requirements, but should also endeavor to be versatile enough to meet those requirements for all the signals displayed in the graphic.

@@

Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single color. 
We want to know "Which of these coefficients are different from 0?" as well as "Which of these coefficients are different from each other?". 
To answer this question we do a series of t-tests on these estimates. 
All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. 
We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. 
Next, we conduct two sample t-test and find that several of the values need to be visually distinguishable from each other. 
The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. 
We cannot perform signal-suppression on both these signals simultaneously.

This example highlights a fundamental problem with the VSUP that extends to the bivariate map as well. 
When we blend these colors, we need to decide at what level of *uncertainty* to blend these colors together. 
Even though the bivariate map does not explicitly combine color values at certain variance levels, the mapping of variance to color saturation does this implicitly. 
That is, at certain saturation values the colors in a bivariate map are imperceptibly different and appear as though they are mapped to the same value.
At this point, it is irrelevant whether or not the colours are technically different, they are the same color in the human brain. 
<!-- This is of course complicated by the fact that human colour perception varies at an individual level: some individuals have been documented to have 4 different cones and can thus perceive a different range of colors; others have only two or one functioning cone and have colour deficiencies. For VSUP to function for all individuals, we must calibrate each plot to an individual's ability to perceive colour.  -->

@@

This does not mean that visualizing raw data will always prevent insignificant signal from getting through. 
@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. 
However, just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
<!-- Does showing the data actually show uncertainty, or does it just show variance? I'm thinking of inferential statistics here - you can show a sample, but it doesn't necessarily show you anything about the uncertainty of the population-based inferential statistics, visually speaking. Some of the dotmap style visualizations you're thinking of are actually census visualizations - there isn't any inferential uncertainty, but there is variance. I'm not sure you're adequately separating these concepts here, even if I generally agree with the broad points you're making.  -->

@@

### Value extraction of uncertainty statistics
Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and the variance [@Kinkeldey2014]. 
This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition. 
By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics. 
This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.

@@

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018]. 
Science communication should be primarily concerned with accuracy. 
<!-- Ok, but you can be completely accurate in how you render a plot and still convey information that isn't accurate (e.g. sine illusion/line width issues). Do we care about accuracy of the visual form, or accuracy of the perception? Because accuracy of the perception is potentially influenced by trust/transparency and potentially influences that metric as well. Rachel Rogers and I have studied how much weight people give testimony in a courtroom - it strikes me that trust/transparency are a separate dimension that is important and related to accuracy, in that both accuracy and trust/transparency will influence how much mental weight I ascribe to new information gained from a visualization.  -->

@@

Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019]. 
Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. 
Risk aversion is considered irrational behaviour because it is a deviation from the behaviour of a rational agent. 
Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational].
However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. 
Rational agents *by definition* should *ignore* uncertainty information. 
Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information.
The only cases where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. 
Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.

@@

The first method is what we call the "vague question" approach. 
The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. 
The authors then go on to compare the readers answer to a very specific ground truth. 
This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. 
Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. 
For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right". 
The mean of the two groups was used to decide which group was actually "furthest to the right" and the participants were evaluated against that ground truth. 
<!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy. Lineup prompts are typically vague (sometimes, I agree, too vauge) because the participant has to figure out what metric to use to evaluate the data first, and then complete the evaluation task. We're as interested in the metric they're using as we are in their successful identification of the target plot. -->
In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. 
This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
<!-- Might be worth referencing the fact that visual statistics and the computational equivalents often differ, and this is even before you start talking measurement error when using a Likert scale.  -->
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts. The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance... and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :). These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm. Land topology also shapes flood risk, which is yet another complexity. All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information. -->
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
Interestingly, it is unclear how the participants were supposed to answer this question. 
@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). 
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.
